# Assignment 3 — Part B

## Setup

```{r setup}
source("00_setup.R", encoding = "UTF-8")
```

## B1. Data exploration

### Read and recode the data

```{r read-and-recode}
raw <- read.csv("data/carrier_data.csv")

d <- raw

d$Y <- factor(
  raw$Y,
  levels = c(1, 2, 3),
  labels = c("A", "B", "C")
)

d$X1 <- factor(
  raw$X1, 
  levels = c(0, 1), 
  labels = c("low", "high")
)

d$X2 <- factor(
  raw$X2, 
  levels = c(0, 1), 
  labels = c("low", "high")
)
d$X3 <- factor(
  raw$X3, 
  levels = c(0, 1), 
  labels = c("low", "high")
)

head(d)
```

### Four-way contingency table

```{r four-way-display}
four_way <- xtabs(~ Y + X1 + X2 + X3, data = d)
ftable(d$X1, d$X2, d$X3, d$Y,
       dnn = c("X1", "X2", "X3", "Y"))
```

### Cross-tabulations of Y with each X

```{r cross-tabulations}
y_x1 <- table(Y = d$Y, X1 = d$X1)
y_x2 <- table(Y = d$Y, X2 = d$X2)
y_x3 <- table(Y = d$Y, X3 = d$X3)

y_x1
y_x2
y_x3
```

### Which attribute appears most related to the preferred provider?

```{r cross-tabulation-proportions}
model_1 <- glm(I(Y == "A") ~ X1 + X2 + X3,
               data = d, family = binomial())

model_2 <- glm(I(Y == "B") ~ X1 + X2 + X3,
               data = d, family = binomial())

model_3 <- glm(I(Y == "C") ~ X1 + X2 + X3,
               data = d, family = binomial())

# Compare coefficients across the three models
round(cbind(
  A = coef(model_1),
  B = coef(model_2),
  C = coef(model_3)
), 3)
```

## B2. Naive Bayes by hand and in software

### B2.1 Naive Bayes DAG and factorization

```{r naive-bayes-dag, fig.width=6, fig.height=4}
nb <- naive.bayes(d, training = "Y")
plot(nb)
```

The model assumes that X1, X2, and X3 are mutually independent conditional on Y.

The joint distribution factorizes as:

$$
p(y,x_1,x_2,x_3)
=
p(y)\,p(x_1\mid y)\,p(x_2\mid y)\,p(x_3\mid y).
$$

### B2.2 Dirichlet priors and posteriors

With global precision $\alpha=12$, the prior parameters are

$$
\alpha_{ijk}=\frac{\alpha}{C_iQ_i}.
$$

For $Y$, we have $C_1=3$ and $Q_1=1$, so

$$
\boldsymbol{\theta}_1 \sim \operatorname{Dir}(4,4,4).
$$

For each attribute, $C_i=2$ and $Q_i=3$. Therefore,
for $j\in\{A,B,C\}$,

$$
\boldsymbol{\theta}_{2j},
\boldsymbol{\theta}_{3j},
\boldsymbol{\theta}_{4j}
\sim \operatorname{Dir}(2,2).
$$

#### Posterior distributions of all parameters

For each parameter vector, Dirichlet–multinomial conjugacy gives

$$
\boldsymbol{\theta}_{ij}\mid D
\sim \operatorname{Dir}
\left(\alpha_{ij1}+N_{ij1},\ldots,
      \alpha_{ijC_i}+N_{ijC_i}\right),
$$

where $N_{ijk}$ is the observed count for category $k$ under
parent configuration $j$. Thus, each posterior parameter is
the corresponding prior parameter plus the observed count.

##### Class probabilities

The category order is $(A,B,C)$. The observed class counts are
$(333,380,164)$. With prior $\operatorname{Dir}(4,4,4)$,

$$
\begin{aligned}
\boldsymbol{\theta}_1\mid D
&\sim \operatorname{Dir}(4+333,\;4+380,\;4+164)\\
&=\operatorname{Dir}(337,\;384,\;168).
\end{aligned}
$$

##### Reputation satisfaction: $X_1$

For each provider, the category order is $(\mathrm{low},\mathrm{high})$
and the prior is $\operatorname{Dir}(2,2)$.
The observed counts for providers A, B, and C are
$(203,130)$, $(133,247)$, and $(112,52)$, respectively.

$$
\begin{aligned}
\boldsymbol{\theta}_{2A}\mid D
&\sim \operatorname{Dir}(2+203,\;2+130)
 = \operatorname{Dir}(205,\;132),\\
\boldsymbol{\theta}_{2B}\mid D
&\sim \operatorname{Dir}(2+133,\;2+247)
 = \operatorname{Dir}(135,\;249),\\
\boldsymbol{\theta}_{2C}\mid D
&\sim \operatorname{Dir}(2+112,\;2+52)
 = \operatorname{Dir}(114,\;54).
\end{aligned}
$$

##### Price satisfaction: $X_2$

The category order is $(\mathrm{low},\mathrm{high})$.
The observed counts for providers A, B, and C are
$(196,137)$, $(242,138)$, and $(104,60)$, respectively.
Adding the prior parameters $(2,2)$ gives

$$
\begin{aligned}
\boldsymbol{\theta}_{3A}\mid D
&\sim \operatorname{Dir}(2+196,\;2+137)
 = \operatorname{Dir}(198,\;139),\\
\boldsymbol{\theta}_{3B}\mid D
&\sim \operatorname{Dir}(2+242,\;2+138)
 = \operatorname{Dir}(244,\;140),\\
\boldsymbol{\theta}_{3C}\mid D
&\sim \operatorname{Dir}(2+104,\;2+60)
 = \operatorname{Dir}(106,\;62).
\end{aligned}
$$

##### Service variety satisfaction: $X_3$

The category order is $(\mathrm{low},\mathrm{high})$.
The observed counts for providers A, B, and C are
$(161,172)$, $(174,206)$, and $(91,73)$, respectively.
Adding the prior parameters $(2,2)$ gives

$$
\begin{aligned}
\boldsymbol{\theta}_{4A}\mid D
&\sim \operatorname{Dir}(2+161,\;2+172)
 = \operatorname{Dir}(163,\;174),\\
\boldsymbol{\theta}_{4B}\mid D
&\sim \operatorname{Dir}(2+174,\;2+206)
 = \operatorname{Dir}(176,\;208),\\
\boldsymbol{\theta}_{4C}\mid D
&\sim \operatorname{Dir}(2+91,\;2+73)
 = \operatorname{Dir}(93,\;75).
\end{aligned}
$$

Under the assumed independent Dirichlet priors and the naive Bayes
likelihood, these ten parameter vectors remain mutually independent
in the posterior.

### B2.3 Posterior classification probabilities by hand

For a customer with
$X_1=\mathrm{high}$, $X_2=\mathrm{low}$, and $X_3=\mathrm{high}$,
we want to calculate

$$
P(Y=j\mid X_1=\mathrm{high},X_2=\mathrm{low},
X_3=\mathrm{high},D),
\qquad j\in\{A,B,C\}.
$$

Under naive Bayes, the unnormalized weight for class $j$ is

$$
w_j =
P(Y=j\mid D)
P(X_1=\mathrm{high}\mid Y=j,D)
P(X_2=\mathrm{low}\mid Y=j,D)
P(X_3=\mathrm{high}\mid Y=j,D).
$$

Each probability is obtained from the corresponding Dirichlet
posterior mean: divide its category parameter by the sum of
the parameters in that vector.

#### Step 1: Obtain the required posterior means

For the class probabilities, the posterior is
$\operatorname{Dir}(337,384,168)$, with total $889$.

For the attributes, use the high component for $X_1$ and $X_3$,
and the low component for $X_2$.

| Class | $P(Y=j\mid D)$ | $P(X_1=\mathrm{high}\mid j,D)$ | $P(X_2=\mathrm{low}\mid j,D)$ | $P(X_3=\mathrm{high}\mid j,D)$ |
| --- | --- | --- | --- | --- |
| A | $337/889$ | $132/337$ | $198/337$ | $174/337$ |
| B | $384/889$ | $249/384$ | $244/384$ | $208/384$ |
| C | $168/889$ | $54/168$ | $106/168$ | $75/168$ |

For example, the posterior for $X_1$ given $Y=A$ is
$\operatorname{Dir}(205,132)$, so

$$
P(X_1=\mathrm{high}\mid Y=A,D)
=\frac{132}{205+132}
=\frac{132}{337}.
$$

#### Step 2: Calculate the unnormalized class weights

$$
\begin{aligned}
w_A
&=\frac{337}{889}
  \frac{132}{337}
  \frac{198}{337}
  \frac{174}{337},\\[4pt]
w_B
&=\frac{384}{889}
  \frac{249}{384}
  \frac{244}{384}
  \frac{208}{384},\\[4pt]
w_C
&=\frac{168}{889}
  \frac{54}{168}
  \frac{106}{168}
  \frac{75}{168}.
\end{aligned}
$$

#### Step 3: Normalize the weights

The weights are joint predictive probabilities. To obtain
conditional class probabilities, divide each by their sum:

$$
P(Y=j\mid x,D)=\frac{w_j}{w_A+w_B+w_C}.
$$

This gives

$$
\begin{aligned}
P(Y=A\mid x,D)&\approx 0.284084,\\
P(Y=B\mid x,D)&\approx 0.608006,\\
P(Y=C\mid x,D)&\approx 0.107910.
\end{aligned}
$$

The three probabilities sum to 1. We classify this customer
as preferring **provider B**, which has the highest posterior
classification probability, approximately **60.80%**.

### B2.4 Verification using bnlearn

#### Fit the naive Bayes model

```{r b2-4-fit}
nb <- naive.bayes(d, training = "Y")

nb_fit <- bn.fit(
  nb,
  data = d,
  method = "bayes",
  iss = 12
)

# Conditional probability tables: posterior means
coef(nb_fit)
```

The argument `iss = 12` corresponds to the global precision
$\alpha=12$. The fitted probability tables contain posterior
means, rather than the full Dirichlet posterior distributions.

#### Verify the posterior Dirichlet parameters

For Y, each prior parameter is $12/(3\times1)=4$.
For each attribute conditional on Y, each prior parameter is
$12/(2\times3)=2$. Adding the observed counts gives:

```{r b2-4-posterior-parameters}
post_Y  <- table(d$Y) + 4
post_X1 <- table(Y = d$Y, X1 = d$X1) + 2
post_X2 <- table(Y = d$Y, X2 = d$X2) + 2
post_X3 <- table(Y = d$Y, X3 = d$X3) + 2

post_Y
post_X1
post_X2
post_X3
```

For Y, the entries form one Dirichlet parameter vector.
For each attribute, each row forms a Dirichlet parameter
vector in the order low/high. These reproduce B2.2.

#### Predict the specified customer's preference

```{r b2-4-predict}
customer <- data.frame(
  X1 = factor("high", levels = levels(d$X1)),
  X2 = factor("low",  levels = levels(d$X2)),
  X3 = factor("high", levels = levels(d$X3))
)

prediction <- predict(
  nb_fit,
  data = customer,
  prob = TRUE
)

prediction
software_probs <- attr(prediction, "prob")
round(software_probs, 6)
```

The probabilities are approximately 0.284084 for A,
0.608006 for B, and 0.107910 for C, matching the hand
calculation in B2.3. The predicted provider is B.

#### Verify using cpquery

```{r b2-4-cpquery}
set.seed(20261002)

query_probs <- c(
  A = cpquery(
    nb_fit,
    event = (Y == "A"),
    evidence = (X1 == "high" & X2 == "low" & X3 == "high"),
    method = "ls",
    n = 1000000
  ),
  B = cpquery(
    nb_fit,
    event = (Y == "B"),
    evidence = (X1 == "high" & X2 == "low" & X3 == "high"),
    method = "ls",
    n = 1000000
  ),
  C = cpquery(
    nb_fit,
    event = (Y == "C"),
    evidence = (X1 == "high" & X2 == "low" & X3 == "high"),
    method = "ls",
    n = 1000000
  )
)

comparison <- data.frame(
  Provider = c("A", "B", "C"),
  Predict = as.numeric(software_probs[c("A", "B", "C"), 1]),
  Cpquery = as.numeric(query_probs)
)

knitr::kable(comparison, digits = 6)
```

With `method = "ls"`, `cpquery()` simulates observations
from the fitted network, retains those matching the evidence,
and estimates the fraction belonging to the requested class.

Its results differ slightly from the exact probabilities
because of Monte Carlo sampling error. Without resetting
the seed, repeated calls generally produce different estimates.
Resetting the same seed makes the simulation reproducible;
increasing the number of samples generally reduces the error.

Because A, B, and C are estimated in separate simulation calls,
their estimated probabilities need not sum to exactly 1.

## B3. Prior sensitivity

We refit the naive Bayes model with $\alpha=1,12,100,1000$
and compare the classification probabilities for the customer
with high reputation, low price, and high service satisfaction.

```{r b3-prior-sensitivity}
alpha_values <- c(1, 12, 100, 1000)

sensitivity <- data.frame(
  alpha = alpha_values,
  A = NA_real_,
  B = NA_real_,
  C = NA_real_
)

for (i in seq_along(alpha_values)) {
  fit <- bn.fit(
    nb,
    data = d,
    method = "bayes",
    iss = alpha_values[i]
  )

  pred <- predict(fit, data = customer, prob = TRUE)

  sensitivity[i, c("A", "B", "C")] <-
    as.numeric(attr(pred, "prob")[c("A", "B", "C"), 1])
}

knitr::kable(sensitivity, digits = 6)
```

The results are:

| $\alpha$ | P(A) | P(B) | P(C) |
| --- | --- | --- | --- |
| 1 | 0.283427 | 0.611108 | 0.105465 |
| 12 | 0.284084 | 0.608006 | 0.107910 |
| 100 | 0.288833 | 0.585496 | 0.125672 |
| 1000 | 0.311772 | 0.470356 | 0.217872 |

The predicted provider remains **B** for all four prior strengths.
However, its classification probability decreases as $\alpha$
increases, while the probabilities of A and C increase.

To explain this, the posterior mean can be written as

$$
E(\theta_{ijk}\mid D)
=
\frac{N_{ijk}+\alpha/(C_iQ_i)}
     {N_{ij}+\alpha/Q_i}
=
\frac{N_{ij}}{N_{ij}+\alpha/Q_i}
\frac{N_{ijk}}{N_{ij}}
+
\frac{\alpha/Q_i}{N_{ij}+\alpha/Q_i}
\frac{1}{C_i}.
$$

Thus, the posterior mean is a weighted average of the observed
proportion and the uniform prior mean $1/C_i$.

When $\alpha$ is small, the data receive more weight.
As $\alpha$ increases, the uniform prior receives more weight,
pulling the conditional probabilities toward uniformity.

With the data held fixed, as $\alpha\to\infty$, the class
probabilities approach $1/3$ and each binary attribute's
conditional probabilities approach $1/2$. Consequently,
the customer's classification probabilities also approach
$(1/3,1/3,1/3)$.

## B4. Five candidate classifiers

### B4.1 Construct the networks



#### Joint factorizations

$$
\begin{aligned}
M_1:\quad p(y,x_1,x_2,x_3)
&=p(y)p(x_1\mid y)p(x_2\mid y,x_1)p(x_3\mid y,x_1,x_2),\\
M_2:\quad p(y,x_1,x_2,x_3)
&=p(y)p(x_1\mid y)p(x_2\mid y)p(x_3\mid y),\\
M_3:\quad p(y,x_1,x_2,x_3)
&=p(y)p(x_1\mid y)p(x_2\mid y,x_1)p(x_3\mid y),\\
M_4:\quad p(y,x_1,x_2,x_3)
&=p(y)p(x_1\mid y)p(x_2\mid y)p(x_3\mid y,x_1),\\
M_5:\quad p(y,x_1,x_2,x_3)
&=p(y)p(x_1\mid y)p(x_2\mid y)p(x_3\mid y,x_2).
\end{aligned}
$$

#### Parent configurations

Let $L=\mathrm{low}$ and $H=\mathrm{high}$.
The parent sets for each model are:

| Model | Parents of Y | Parents of X1 | Parents of X2 | Parents of X3 |
| --- | --- | --- | --- | --- |
| M1 | None | Y | (Y, X1) | (Y, X1, X2) |
| M2 | None | Y | Y | Y |
| M3 | None | Y | (Y, X1) | Y |
| M4 | None | Y | Y | (Y, X1) |
| M5 | None | Y | Y | (Y, X2) |

Each parent set has the following configurations.
Tuple entries follow the parent order shown above.

- **No parents:** one empty configuration, so $Q=1$.
- **Y:** $A,B,C$, so $Q=3$.
- **(Y, X1):**
  $(A,L),(A,H),(B,L),(B,H),(C,L),(C,H)$, so $Q=6$.
- **(Y, X2):**
  $(A,L),(A,H),(B,L),(B,H),(C,L),(C,H)$, so $Q=6$.
- **(Y, X1, X2):** the following twelve configurations, so $Q=12$:

$$
\begin{aligned}
&(A,L,L),\ (A,L,H),\ (A,H,L),\ (A,H,H),\\
&(B,L,L),\ (B,L,H),\ (B,H,L),\ (B,H,H),\\
&(C,L,L),\ (C,L,H),\ (C,H,L),\ (C,H,H).
\end{aligned}
$$

#### B4.2(a) Parameter estimation and Dirichlet posteriors

#### Define and fit the five models

```{r b4-models}
M1 <- model2network("[Y][X1|Y][X2|Y:X1][X3|Y:X1:X2]")
M2 <- model2network("[Y][X1|Y][X2|Y][X3|Y]")
M3 <- model2network("[Y][X1|Y][X2|Y:X1][X3|Y]")
M4 <- model2network("[Y][X1|Y][X2|Y][X3|Y:X1]")
M5 <- model2network("[Y][X1|Y][X2|Y][X3|Y:X2]")

fit1 <- bn.fit(M1, data = d, method = "bayes", iss = 12)
fit2 <- bn.fit(M2, data = d, method = "bayes", iss = 12)
fit3 <- bn.fit(M3, data = d, method = "bayes", iss = 12)
fit4 <- bn.fit(M4, data = d, method = "bayes", iss = 12)
fit5 <- bn.fit(M5, data = d, method = "bayes", iss = 12)
```

```{r b4-plots, fig.width=9, fig.height=6}
par(mfrow = c(2, 3))

plot(M1, main = "M1")
plot(M2, main = "M2")
plot(M3, main = "M3")
plot(M4, main = "M4")
plot(M5, main = "M5")

par(mfrow = c(1, 1))
```

We report the complete Dirichlet posterior distributions
for M3, whose factorization is

$$
p(y,x_1,x_2,x_3)
=
p(y)p(x_1\mid y)p(x_2\mid y,x_1)p(x_3\mid y).
$$

With $\alpha=12$, the prior parameters are

$$
\alpha_{ijk}=\frac{12}{C_iQ_i}.
$$

Therefore:

- Y has prior Dir(4,4,4).
- Each X1 given Y vector has prior Dir(2,2).
- Each X2 given (Y,X1) vector has prior Dir(1,1).
- Each X3 given Y vector has prior Dir(2,2).

The posterior parameters equal the prior parameters plus
the corresponding observed counts.

##### Class probabilities: Y

The category order is A, B, C.

$$
\boldsymbol{\theta}_1\mid D
\sim \operatorname{Dir}(4+333,\;4+380,\;4+164)
=\operatorname{Dir}(337,\;384,\;168).
$$

##### X1 conditional on Y

Each vector is ordered low, high.

$$
\begin{aligned}
\boldsymbol{\theta}_{2,A}\mid D
&\sim \operatorname{Dir}(2+203,\;2+130)
=\operatorname{Dir}(205,\;132),\\
\boldsymbol{\theta}_{2,B}\mid D
&\sim \operatorname{Dir}(2+133,\;2+247)
=\operatorname{Dir}(135,\;249),\\
\boldsymbol{\theta}_{2,C}\mid D
&\sim \operatorname{Dir}(2+112,\;2+52)
=\operatorname{Dir}(114,\;54).
\end{aligned}
$$

##### X2 conditional on Y and X1

There are six parent configurations, so each prior parameter
is $12/(2\times6)=1$.

In the subscripts below, $(A,L)$ means $Y=A$ and
$X_1=\mathrm{low}$. Each parameter vector describes the
probabilities of X2 in the order low, high.

$$
\begin{aligned}
\boldsymbol{\theta}_{3,(A,L)}\mid D
&\sim \operatorname{Dir}(1+148,\;1+55)
=\operatorname{Dir}(149,\;56),\\
\boldsymbol{\theta}_{3,(A,H)}\mid D
&\sim \operatorname{Dir}(1+48,\;1+82)
=\operatorname{Dir}(49,\;83),\\
\boldsymbol{\theta}_{3,(B,L)}\mid D
&\sim \operatorname{Dir}(1+116,\;1+17)
=\operatorname{Dir}(117,\;18),\\
\boldsymbol{\theta}_{3,(B,H)}\mid D
&\sim \operatorname{Dir}(1+126,\;1+121)
=\operatorname{Dir}(127,\;122),\\
\boldsymbol{\theta}_{3,(C,L)}\mid D
&\sim \operatorname{Dir}(1+90,\;1+22)
=\operatorname{Dir}(91,\;23),\\
\boldsymbol{\theta}_{3,(C,H)}\mid D
&\sim \operatorname{Dir}(1+14,\;1+38)
=\operatorname{Dir}(15,\;39).
\end{aligned}
$$

##### X3 conditional on Y

Each vector is ordered low, high.

$$
\begin{aligned}
\boldsymbol{\theta}_{4,A}\mid D
&\sim \operatorname{Dir}(2+161,\;2+172)
=\operatorname{Dir}(163,\;174),\\
\boldsymbol{\theta}_{4,B}\mid D
&\sim \operatorname{Dir}(2+174,\;2+206)
=\operatorname{Dir}(176,\;208),\\
\boldsymbol{\theta}_{4,C}\mid D
&\sim \operatorname{Dir}(2+91,\;2+73)
=\operatorname{Dir}(93,\;75).
\end{aligned}
$$

These are all 13 parameter vectors for M3:
one for Y, three for X1, six for X2, and three for X3.

#### B4.2(b) Prediction methods and classification probabilities

##### Create the eight profiles

Writing them explicitly makes their order easy to follow:

```{r b4-profiles}
profiles <- data.frame(
  X1 = c("low", "low", "low", "low",
         "high", "high", "high", "high"),

  X2 = c("low", "low", "high", "high",
         "low", "low", "high", "high"),

  X3 = c("low", "high", "low", "high",
         "low", "high", "low", "high")
)

profiles$X1 <- factor(profiles$X1, levels = levels(d$X1))
profiles$X2 <- factor(profiles$X2, levels = levels(d$X2))
profiles$X3 <- factor(profiles$X3, levels = levels(d$X3))

knitr::kable(profiles)
```

##### Compare the three prediction methods

Here is one small helper function. It takes a fitted model and returns all three methods’ probabilities for the eight profiles.

```{r b4-methods}
compare_methods <- function(fitted_model) {

  pred_parents <- predict(
    fitted_model, node = "Y", data = profiles,
    method = "parents", prob = TRUE
  )

  set.seed(20261002)
  pred_lw <- predict(
    fitted_model, node = "Y", data = profiles,
    method = "bayes-lw", prob = TRUE, n = 100000
  )

  pred_exact <- predict(
    fitted_model, node = "Y", data = profiles,
    method = "exact", prob = TRUE
  )

  # Probability matrices have classes in rows.
  # t() transposes them so each customer profile becomes a row.
  p_parents <- t(attr(pred_parents, "prob")[c("A", "B", "C"), ])
  p_lw <- t(attr(pred_lw, "prob")[c("A", "B", "C"), ])
  p_exact <- t(attr(pred_exact, "prob")[c("A", "B", "C"), ])

  result <- rbind(
    data.frame(Method = "parents", profiles, p_parents),
    data.frame(Method = "bayes-lw", profiles, p_lw),
    data.frame(Method = "exact", profiles, p_exact)
  )

  return(result)
}

methods1 <- compare_methods(fit1)
methods2 <- compare_methods(fit2)
methods3 <- compare_methods(fit3)
methods4 <- compare_methods(fit4)
methods5 <- compare_methods(fit5)

knitr::kable(methods1, digits = 4, caption = "M1")
knitr::kable(methods2, digits = 4, caption = "M2")
knitr::kable(methods3, digits = 4, caption = "M3")
knitr::kable(methods4, digits = 4, caption = "M4")
knitr::kable(methods5, digits = 4, caption = "M5")
```

##### Exact probabilities and predicted classes

This function selects the largest probability in each row. If classes are tied within numerical tolerance, it chooses A before B before C.

```{r b4-exact}
get_results <- function(fitted_model, new_data) {

  prediction <- predict(
    fitted_model, node = "Y", data = new_data,
    method = "exact", prob = TRUE
  )

  probabilities <- t(
    attr(prediction, "prob")[c("A", "B", "C"), , drop = FALSE]
  )

  predicted_class <- character(nrow(new_data))
  classes <- c("A", "B", "C")

  for (i in seq_len(nrow(new_data))) {
    highest <- max(probabilities[i, ])

    # Find classes tied for the highest probability.
    candidates <- which(probabilities[i, ] >= highest - 1e-12)

    # Choose the first: A, then B, then C.
    predicted_class[i] <- classes[candidates[1]]
  }

  result <- data.frame(
    new_data,
    probabilities,
    Predicted = predicted_class
  )

  return(result)
}

result1 <- get_results(fit1, profiles)
result2 <- get_results(fit2, profiles)
result3 <- get_results(fit3, profiles)
result4 <- get_results(fit4, profiles)
result5 <- get_results(fit5, profiles)

# Each cell shows (A, B, C) probabilities and the prediction.
format_result <- function(result) {
  sprintf(
    "(%.4f, %.4f, %.4f) -> %s",
    result$A, result$B, result$C, result$Predicted
  )
}

comparison <- data.frame(
  profiles,
  M1 = format_result(result1),
  M2 = format_result(result2),
  M3 = format_result(result3),
  M4 = format_result(result4),
  M5 = format_result(result5)
)

knitr::kable(comparison)
```

#### B4.2(c) Confusion matrices and accuracy

We use the same tie rule when predicting the original observations. These are **in-sample** results.

```{r b4-accuracy}
# Only attributes are supplied as prediction inputs.
attributes <- d[, c("X1", "X2", "X3")]

training1 <- get_results(fit1, attributes)
training2 <- get_results(fit2, attributes)
training3 <- get_results(fit3, attributes)
training4 <- get_results(fit4, attributes)
training5 <- get_results(fit5, attributes)

# Keep all three prediction categories, even if one is never selected.
pred1 <- factor(training1$Predicted, levels = c("A", "B", "C"))
pred2 <- factor(training2$Predicted, levels = c("A", "B", "C"))
pred3 <- factor(training3$Predicted, levels = c("A", "B", "C"))
pred4 <- factor(training4$Predicted, levels = c("A", "B", "C"))
pred5 <- factor(training5$Predicted, levels = c("A", "B", "C"))

table(Actual = d$Y, Predicted = pred1)
table(Actual = d$Y, Predicted = pred2)
table(Actual = d$Y, Predicted = pred3)
table(Actual = d$Y, Predicted = pred4)
table(Actual = d$Y, Predicted = pred5)

accuracy1 <- mean(pred1 == d$Y)
accuracy2 <- mean(pred2 == d$Y)
accuracy3 <- mean(pred3 == d$Y)
accuracy4 <- mean(pred4 == d$Y)
accuracy5 <- mean(pred5 == d$Y)

accuracy_table <- data.frame(
  Model = c("M1", "M2", "M3", "M4", "M5"),
  Accuracy = c(accuracy1, accuracy2, accuracy3, accuracy4, accuracy5)
)

knitr::kable(accuracy_table, digits = 6)
```

### B4.3 Comparison by log marginal likelihood

#### Marginal likelihood and posterior model probabilities

```{r b4-model-comparison}
score1 <- score(M1, data = d, type = "bde", iss = 12)
score2 <- score(M2, data = d, type = "bde", iss = 12)
score3 <- score(M3, data = d, type = "bde", iss = 12)
score4 <- score(M4, data = d, type = "bde", iss = 12)
score5 <- score(M5, data = d, type = "bde", iss = 12)

log_scores <- c(score1, score2, score3, score4, score5)

# Equal model priors cancel when we normalize.
weights <- exp(log_scores - max(log_scores))
model_probabilities <- weights / sum(weights)

model_comparison <- data.frame(
  Model = c("M1", "M2", "M3", "M4", "M5"),
  Parameters = c(
    nparams(fit1), nparams(fit2), nparams(fit3),
    nparams(fit4), nparams(fit5)
  ),
  Log_marginal_likelihood = log_scores,
  Posterior_probability = formatC(
    model_probabilities, format = "e", digits = 3
  ),
  Accuracy = c(accuracy1, accuracy2, accuracy3, accuracy4, accuracy5)
)

knitr::kable(model_comparison, digits = 6)

# Strength of evidence for M1 over M5, the second-best model
score1 - score5
exp(score1 - score5)
```

#### Verify the contribution of Y by hand

```{r b4-hand-check}
# Prior: Dir(4, 4, 4)
# Counts: (333, 380, 164)
# Posterior: Dir(337, 384, 168)

manual_Y <- lgamma(12) - lgamma(889) +
  lgamma(337) + lgamma(384) + lgamma(168) -
  3 * lgamma(4)

node_scores <- score(
  M2, data = d, type = "bde", iss = 12, by.node = TRUE
)

manual_Y
node_scores["Y"]

# The sum of node contributions equals the whole-network score.
sum(node_scores)
score2
```

Both Y calculations should give approximately **−920.242329**. The overall conclusion stays the same: **M1 has the strongest model evidence, while all five models have 51.31% in-sample accuracy under the stated tie rule.**
